Texture mapping 3D planar models of indoor environments with noisy camera poses

### 2013-01-20 spie/paper.tex
just geom figure now

As Figure \ref{fig:compareAll}(a) demonstrates, this approach seleccts
images with high resolution, and handles parallax effects well, as
seen with the table and the poster at the lower-right of the
image. Obviously, there are many image boundaries with abrupt
discontinuities between tiles, due to significant misalignment between
images, though most of it appears to be reconcilable using 2d
transforms. Given our high number of input images, a better image
selection procedure will further improve results, as described in
Section \ref{sec:imageCompositing}, but a more significant and
reliable improvement can be found through first improving the
alignment of our images.

@@

  \caption{Texture alignment via (a) image mosaicing, (b) the
    graph-based localization refinement algorithm from
    \cite{chen2010indoor}, and (c) the method proposed in this paper.}
  \label{fig:mosaic3D}
\end{figure}

Past work with our data acquisition backpack integrated image
stitching with an iterative localization algorithm, and performed
homography-based camera pose refinement. When run on long chains of
images, especially where features are sparse, it produces distorted
textures, as seen in Figure \ref{fig:mosaic3D}(a), due to the problems
described above. This approach is also not closed-form, and its
iterative camera adjustment process over our large datasets leads to
prohibitively long computation time.

@@

\section{2D Image Alignment with Geometry Information}
\label{sec:2dAlignment}

In this section, we describe our method of efficient and robust image
alignment. Though state-of-the-art techniques towards image stitching
work in 3D, we elect to perform image alignments entirely in
2D. Besides the benifits of reduced complexity, we have found 2D
alignment to work well for two main reasons. First, because we have
such a large number of images to choose from, the compositing
approaches in Section \ref{sec:imageCompositing} are able to select a
subset of images that align well, given only 2D alignments. Second,
the nature of our input data is such that localization error chiefly
occurs in two dimensions, in the same plane as the surface being
projected onto. Our data acquisition system contains cameras facing to
the sides of the operator. The operator makes efforts to walk parallel
to walls, and the localization and model-generation algorithms that
provide our input conform the operator's path to be straight and
parallel to detected surfaces \cite{kua2012loopclosure,
  sanchez2012point}. Since the operator walks in as upright a position
as possible, errors in roll, yaw, and the operator's distance from
parallel surfaces are minimal. Thus, our highest errors stem from
uncertainty in the operator's pitch, equivalent to rotation around the
camera axes, as well as location along surfaces. These equate to 2D
rotation and translation in a surface's plane.

Our 2D alignment procedure consists of three main parts. First, all
images are projected onto the surface and lines within these
projections are detected. These lines are then rotated and shifted to
match geometry-based lines comprising the surface's boundary and
intersection with other surfaces. Second, occlusion checks are
performed to remove invalid parts of each image for the target
surface. These two steps are image-independent and performed in
parallel. Third, we detect SIFT feature matches between pairs of
images and solve a weighted linear least squares problem in 2D to
maximize matches.

@@

To perform rotation, we use a RANSAC \cite{fischler1981random}
framework to determine an optimal rotation angle between lines in our
image and lines in our geometry. RANSAC allows us to ignore erroneous
lines in our image, such as the slanted lines corresponding to the
ladder in Figure \ref{fig:geometryAlignment}(a). Once this rotation is
discovered and applied, we gather all pairs of image line segments and
geometry line segments with less than $0.1^\circ$ difference in
orientation and less than 250 mm distance at their furthest points. If
2 or more pairs exist, we select the 2 that contain the longest
noncollinear image-based line segments, and perform the corresponding
fixed translation to match them. If we only have 1 pair, we perform
the minimal shift to match the lines of that pair, and make note of
their orientation, for usage in Section
\ref{robustSIFTFeatureMatching}.

\subsection{Image Occlusion}
\label{sec:imageOcclusion}
Now that images have been aligned to geometry where possible, we
perform a simple recursive occlusion procedure on each image. This is
done by performing the intersection tests in section
\ref{sec:simpleTextureMapping} in a regularly spaced grid. Where four
corners of a rectangular region are occluded, texture is
removed. Where no corners are occluded, the recursion stops. Where
there is a mixture of both, the rectangular region is subdivided into
four, and the same process is performed on each. By performing Section
\ref{sec:geometryAlignment}'s alignment procedure before occlusion,
texture belonging to other surfaces is accurately removed, which is
necessary for the next section.

\subsection{2D Feature Alignment}
\label{sec:robustSIFTFeatureMatching}
Our next step is to align overlapping images by searching for
corresponding points between all pairs of overlapping images. We use
SIFT features for their high detection rate, and choose to use feature
alignment rather than pixel or intensity-based alignment due to the
differences in lighting as well as possible occlusion among our
images, both of which feature alignment is less sensitive to
\cite{lowe1999object, mikolajczyk2005performance, szeliski2006image}.
We use SiftGPU \cite{siftgpu} for its high performance on both feature
detection as well as pairwise matching. These matches determine $d^x$
and $d^y$ distances between each pair of features for two image
projections, though these distances may not always be the same for
different features. Since indoor environments often contain repetitive
features such as floor tiles or doors, we need to ensure that
SIFT-based distances are reliable. First, we only perform alignment on
parts of images that overlap given the original noisy poses. Second,
we discard feature matches that correspond to an image distance
greater than 40 pixels from what the noisy poses estimate. In order to
utilize the remaining feature matches robustly, RANSAC
\cite{fischler1981random} is again used to estimate the optimal
$d^x_{i,j}$ and $d^y_{i,j}$ distances between two images $i$ and
$j$. We use a 10 pixel threshold for RANSAC, so that SIFT matches are
labeled as outliers if their distance is not within 10 pixels of the
sampled average distance.

We now use the $d^x_{i,j}$ and $d^y_{i,j}$ distances between each pair
of images to refine their positions using weighted linear least
squares. The variables we wish to solve for are the $x$ and $y$
positions of our images, while our equations are the SIFT-based
distances between pairs of images, images fixed to geometry with 0 or
1 degrees of freedom, and the original camera poses. Ignoring
weighting for now, an example setup for solving
$\textrm{min}_{\vec{\beta}} ||A \vec{\beta} - \vec{\gamma}||_2^2 $
with 3 images is shown below.

@@

In this scenario, a SIFT-based distance of $dx_{1,2}$, $dy_{1,2}$ was
calculated between images 1 and 2. This corresponds to the first and
second of $A$, while the third and fourth row of $A$ represent the
same for images 2 and 3. Rows 5 and 6 of $A$ correspond to a fixed
location of $tx_1$, $ty_1$ for image 1, while Row 7 corresponds to a
constraint to a line of slope $m$, with current location $tx_2$,
$ty_2$, both results of the geometry alignment procedure in Section
\ref{sec:geometryAlignment}. If we do not have enough SIFT matches, or
lack images matched to geometry, our matrix becomes rank-deficient and
our problem cannot be solved. As a result, we add rows for each image
corresponding to the original noisy image locations, and downweight
them heavily, e.g. with a weight of 0.01 versus 1.

Because our problem is linear, it is easily and quickly solved, and after
applying the resulting shifts, our images overlap and match each other
with far greater accuracy. Applying the simple mapping scheme in
Section \ref{sec:tileBasedMapping} to the same wall used in that
section results in Figure \ref{fig:compareAll}(b), which has far fewer
discontinuities, though errors due to lighting differences and
repeating features are still visible.

@@

The result of this caching approach is shown in Figure
\ref{fig:compareAll}(c), where seams are now reduced as compared to
Figure \ref{fig:compareAll}(b). However, some discontinuities are
still present, as visible in the many posters on the wall with breaks
in their borders.

@@

\begin{figure}
  \centering
  \includegraphics[width=3in]{dagCreation.pdf}
  \caption{DAG construction for the image selection process. \\}
  \label{fig:dagCreation}
\end{figure}

@@

As mentioned earlier, our approach is quite efficient. The top wall in
Figure \ref{fig:results}(a) was generated with 7543 $\times$ 776
pixels, and spans a 40-meter long wall. Given 41000 input images, a
2.8GHz consumer-grade laptop takes approximately a minute to pick 36
candidate images, followed by another minute to perform both image
alignment and the shortest path texturing method, though over 75\% of
that time is spent calculating SIFT matches within the SiftGPU
framework. While not real-time, the process is capable of generating
quick updates after changes in various parameters or modifications to
input data, and if integrated directly into a modeling system, could
provide live visualization as data is collected. Our full models
consist of our input model file, textures that we generate, and a
mapping of image points to model vertices. The models shown in Figure
\ref{fig:results} are roughly 20 MB in size, and are visualized using
the OpenSceneGraph toolkit \cite{openscenegraph}, which allows for export to many common
model formats, as well as efficient visualization, even in web
browsers or mobile devices.

>>>>

As Figure \ref{fig:compareAll}(a) demonstrates, there are many image
boundaries with abrupt discontinuities between tiles, due to
significant misalignment between images, though most of it appears to
be reconcilable using 2D transforms. Given our high number of input
images, a better image selection procedure will further improve
results, as described in Section \ref{sec:imageCompositing}, but a
more significant and reliable improvement can be found through first
improving the alignment of our images.

@@

  \caption{Texture alignment via (a) the graph-based localization
    refinement algorithm, (b) the AutoStitch software package, and (c)
    our method.}
  \label{fig:mosaic3D}
\end{figure}

Past work with our data acquisition backpack integrated image
stitching with an iterative localization algorithm, and performed
homography-based camera pose refinement \cite{liu2010indoor}. When run
on long chains of images, especially where features are sparse, it
produces distorted textures, as seen in Figure \ref{fig:mosaic3D}(a),
due to the problems described above. This approach is also not
closed-form, and its iterative camera adjustment process over our
large datasets leads to prohibitively long computation time.

@@

\section{2D Image Alignment}
\label{sec:2dAlignment}

In this section, we describe our method of efficient and robust image
alignment. Though state-of-the-art techniques towards image stitching
work in 3D, we elect to perform image alignments entirely in
2D. Besides the benifits of reduced complexity, we have found 2D
alignment to work well for two main reasons. First, because we have
such a large number of images to choose from, the compositing
approaches in Section \ref{sec:imageCompositing} are able to select a
subset of images that align well, given only 2D alignments. Second,
the nature of our input data is such that localization error chiefly
occurs in two dimensions, in the same plane as the surface being
projected onto. Our data acquisition system contains cameras facing to
the sides of the operator. The operator makes efforts to walk parallel
to walls, and the localization and model-generation algorithms that
provide our input conform the operator's path to be straight and
parallel to detected surfaces \cite{kua2012loopclosure,
  sanchez2012point}. Since the operator also walks in a manner to keep
the backpack balanced, errors in roll and yaw are minimal. Thus, our
highest errors stem from uncertainty in the operator's pitch,
equivalent to rotation around the camera axes, as well as location
along surfaces. These equate to 2D rotation and translation in a
surface's plane.

Our 2D alignment procedure consists of three parts. First, all images
are projected onto the surface and lines within these projections are
detected. These lines are then rotated and shifted to match
geometry-based lines comprising the surface's boundary and
intersection with other surfaces. Second, occlusion checks are
performed to remove invalid parts of each image for the target
surface. These two steps are image-independent and performed in
parallel. Third, we detect SIFT feature matches between pairs of
images and solve a weighted linear least squares problem in 2D to
maximize matches.

@@

To perform rotation, we use a RANSAC \cite{fischler1981random}
framework to determine an optimal rotation angle between lines in our
image and lines in our geometry. RANSAC allows us to ignore erroneous
lines in our image, such as the slanted lines corresponding to the
ladder in Figure \ref{fig:geometryAlignment}(a). Once this rotation is
discovered and applied, we gather all pairs of image line segments and
geometry line segments with less than $0.1^\circ$ difference in
orientation and less than 250 mm distance at their furthest points. If
2 or more pairs exist, we select the 2 that contain the longest
noncollinear image-based line segments, and perform the corresponding
fixed translation to match them. If we only have 1 pair, we perform
the minimal shift to match the lines of that pair, and make note of
their orientation, for usage in Section
\ref{sec:robustSIFTFeatureMatching}.

\subsection{Image Occlusion}
\label{sec:imageOcclusion}
Now that images have been aligned to geometry where possible, we
perform a simple recursive occlusion procedure on each image. This is
done by performing the intersection tests in Section
\ref{sec:simpleTextureMapping} in a regularly spaced grid. Where four
corners of a rectangular region are occluded, texture is
removed. Where no corners are occluded, the recursion stops. Where
there is a mixture of both, the rectangular region is subdivided into
four, and the same process is performed on each. By performing Section
\ref{sec:geometryAlignment}'s alignment procedure before occlusion,
texture belonging to other surfaces is accurately removed, which is
necessary for the next section.

\subsection{2D Feature Alignment}
\label{sec:robustSIFTFeatureMatching}
Our next step is to align images by searching for corresponding points
between all pairs of overlapping images. We use feature alignment
rather than pixel or intensity-based alignment due to the differences
in lighting as well as possible occlusion among our images, both of
which feature alignment is less sensitive to \cite{lowe1999object,
  mikolajczyk2005performance, szeliski2006image}.  We use SiftGPU
\cite{siftgpu} for its high performance on both feature detection as
well as pairwise matching. These matches determine $dx$ and $dy$
distances between each pair of features for two image projections,
though these distances may not always be the same for different
features. Since indoor environments often contain repetitive features
such as floor tiles or doors, we need to ensure that SIFT-based
distances are reliable. First, we only perform alignment on parts of
images that overlap given the original noisy poses. Second, we discard
feature matches that correspond to an image distance greater than 200
mm from what the noisy poses estimate. In order to utilize the
remaining feature matches robustly, RANSAC \cite{fischler1981random}
is again used to estimate the optimal $dx_{i,j}$ and $dy_{i,j}$
distances between two images $i$ and $j$. We use a 5 mm threshold for
RANSAC, so that SIFT matches are labeled as outliers if their distance
is not within 5 mm of the sampled average distance.

We now use the $dx_{i,j}$ and $dy_{i,j}$ distances between each pair
of images to refine their positions using weighted linear least
squares. The variables we wish to solve for are the $_ix$ and $y_i$
positions of our images, while our equations are the SIFT-based
distances between pairs of images, images fixed to geometry with 0 or
1 degrees of freedom, and the original camera poses. Ignoring
weighting and the original camera poses for now, an example setup for
solving $\textrm{min}_{\vec{\beta}} ||A \vec{\beta} -
\vec{\gamma}||_2^2 $ with 3 images is shown below.

@@

In this scenario, a SIFT-based distance of $dx_{1,2}$, $dy_{1,2}$ was
calculated between images 1 and 2. This corresponds to the first and
second row of $A$, while the third and fourth row of $A$ represent the
same for images 2 and 3. Rows 5 and 6 of $A$ correspond to a fixed
location of $tx_1$, $ty_1$ for image 1, while Row 7 corresponds to a
constraint to a line of slope $m$, with current location $tx_2$,
$ty_2$, both possible results of the geometry alignment procedure in
Section \ref{sec:geometryAlignment}. If we do not have enough SIFT
matches, or lack images matched to geometry, our matrix becomes
rank-deficient and our problem cannot be solved. As a result, we add
rows for each image corresponding to the original noisy image
locations, and downweight them heavily, e.g. with a weight of 0.01
versus 1.

Because our problem is linear, it is easily and quickly solved, and
after applying the resulting shifts, our images overlap and match each
other with far greater accuracy. Applying the simple mapping scheme in
Section \ref{sec:tileBasedMapping} to the same wall used in that
section results in Figure \ref{fig:compareAll}(b), which has far fewer
discontinuities, though errors due to lighting differences and
parallax effects are still visible.

@@

The result of this caching approach is shown in Figure
\ref{fig:compareAll}(c), where seams are now reduced as compared to
Figure \ref{fig:compareAll}(b). However, some discontinuities are
still present, as visible in the posters on the wall with breaks
in their borders.

@@

\begin{figure}
  \centering
  \includegraphics[width=5in]{dagCreation.pdf}
  \caption{DAG construction for the image selection process. \\}
  \label{fig:dagCreation}
\end{figure}

@@

As mentioned earlier, our approach is quite efficient. The top wall in
Figure \ref{fig:results}(a) was generated with 7543 $\times$ 776
pixels, and spans a 40-meter long wall. Given 41000 input images, a
2.8GHz dual-core consumer-grade laptop takes approximately a minute to
pick 36 candidate images, followed by another minute to perform both
image alignment and the shortest path texturing method, though over
75\% of that time is spent calculating SIFT matches within the SiftGPU
framework. While not real-time, the process is capable of generating
quick updates after changes in various parameters or modifications to
input data, and if integrated directly into a modeling system, could
provide live visualization as data is collected. Our full models
consist of our input model file, textures that we generate, and a
mapping of image points to model vertices. The models shown in Figure
\ref{fig:results} are roughly 20 MB in size, and are visualized using
the OpenSceneGraph toolkit \cite{openscenegraph}, which allows for
export to many common model formats, as well as efficient
visualization, even in web browsers or mobile devices.
